---
output:
  md_document
date: "2025-05-14"
---

```{r libraries, message=F}

library(neonUtilities)
library(arrow)
library(dplyr)
library(lubridate)
library(ggplot2)

```

The neonUtilities version used here is the development version, not the released version.

## Last time:

* `queryFiles()` is a neonUtilities function, a wrapper for the new API endpoint that returns the full list of files meeting the search criteria. https://data.neonscience.org/data-api/endpoints/data-query/

* `open_csv_dataset()` is an arrow function. It takes a list of filepaths and defines them as a unified dataset, but doesn't download them.

* `schemaFromVar()` is a neonUtilities function that takes a NEON variables file and translates it into an arrow schema, defining the column names and data types for the dataset.

```{r qfsm, results="hide", message=F}

urllst <- queryFiles(dpID='DP1.10072.001', site='TREE',
                      package='basic', release='RELEASE-2025', 
                      tabl='mam_pertrapnight', metadata=F,
                      token=Sys.getenv('NEON_TOKEN'))

ds <- open_csv_dataset(sources=urllst[[1]], 
                       schema=schemaFromVar(urllst[[2]], 
                                            'mam_pertrapnight',
                                            'basic'), 
                       skip=1)

```

## New function: datasetQuery()

This function combines the functions above into a single step.

```{r qds, message=F}

dsq <- datasetQuery(dpID='DP1.10072.001', site='TREE',
                    package='basic', release='RELEASE-2025', 
                    tabl='mam_pertrapnight', 
                    token=Sys.getenv('NEON_TOKEN'))

```

This one function creates an arrow dataset based on the query criteria.

```{r ds-class}

class(dsq)

```

## IS data

Here's what this function looks like for IS data:

```{r dsq-IS, message=F}

dssw <- datasetQuery(dpID='DP1.00094.001', site='SJER',
                     startdate='2024-04', enddate='2024-06',
                     package='basic', release='RELEASE-2025', 
                     tabl='SWS_30_minute', hor='002', ver='501',
                     token=Sys.getenv('NEON_TOKEN'))

```

`site`, `hor`, and `ver` are required inputs, and only allow for a single 
entry each. This prevents users from unknowingly calculating averages and 
other statistics across locations.

`queryFiles()` is still in the package, so people can still figure out how 
to do more complicated and risky queries on their own, and it doesn't change 
the risk of e.g. leaving out critical columns. But making `datasetQuery()` 
the default way to do this reduces some of the biggest risks.



