---
title: "Example workflow for processing NEON eddy-covariance turbulence data with eddy4R-Docker 0.2.0"
author: "David Durden, Stefan Metzger, Natchaya Pingintha-Durden, Claire Lunch, Megan Jones"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example workflow for processing NEON eddy-covariance turbulence data with eddy4R-Docker 0.2.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 0. Install and set up

This vignette assumes you are working with eddy4R within the Docker 
environment established by NEON. To get set up in that environment, 
follow the instructions <a href="https://github.com/NEONScience/NEON-Data-Skills/blob/fbd76ce778bc369fa9b94b7751546abb1ae56073/tutorials/R/eddy4r/dockerEddy4r.md" target="_blank">here</a>.
(CKL: I put a link to the page on GitHub, because I don't think we've 
put this up on the website yet)

At the end of those instructions, you should have an RStudio 
environment running within Docker. Use that workspace for this 
vignette.


## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
